---
title: "Project 1"
author: "Diego Diaz"
date: "September 27, 2015"
output: 
  html_document: 
    toc: yes
---

For this project I will use the stringr library of functions.

#Loading Necessary Libraries

``` {r load_library}

library(stringr)
```

#Retrieving the Data
Data for this project will be retrieved from Github.  
```{r getting_data}

project_url = url("https://raw.githubusercontent.com/diegomdiaz/IS607/master/Project%201/tournamentinfo.txt")

project_data = readLines(project_url, n = 198)

```

#Initial Cleanup

Extracting only digits and words to get rid of lines that are not useful in this analysis.
 
``` {r extracting_words}
mydata <- unlist(str_extract_all(project_data, "\\W{1,2}.+\\w+"))

```

``` {r removing_first_two_rows}
#Removing the first two rows of data since they are not needed.  
mydata <- mydata [-c(1:2)]

```

#Subsetting the Data Using Logical Vectors

At this point, I thought it would be easier if I split the data using a logical vector and then use do my matching. 

``` {r separting_names_ids}

mydata_names <- mydata [c(TRUE, FALSE)]

mydata_ids <- mydata [c(FALSE, TRUE)]

head(mydata_names)

head(mydata_ids)

```

#Extracting Player Names
Extracting the names of the chess players using the str_extract_all function. 

``` {r extracting_name}

mydata_names1 <- unlist(str_extract_all(mydata_names, "(\\b[[:upper:]-]+\\b\\s)+(\\b[[:upper:]-]+\\b){1}"))

#taking a peak at my new vector. 
head(mydata_names1)

```

#Extracting State Information
Extracting the State information for each player using the str_extract_all function. 

``` {r extracting_state}

mydata_state <- unlist(str_extract_all(mydata_ids, "(\\b[[:upper:]]{2}\\b)"))

#taking a peak at my new vector.
head(mydata_state)

```

#Extracting Total Points
Extracting Total Point information for each player using the str_extract_all function.

``` {r extracting_tota_points}

mydata_total <- unlist(str_extract_all(mydata_names, "[[:digit:]]\\.[[:digit:]]"))

#taking a peak at my new vector.                            
head(mydata_total)

```

#Extracting Pre Scores

Extracting the Pre Score for each player in two steps using the str_extract_all function.

``` {r extract_prescore }

#Looking to match a pattern where we have R: followed by a space and digits. The is followed by a quantifier to match it at least one time, but not more than five times. No value should exceed this limit.

pre_score <- unlist(str_extract_all(mydata_ids, "R:[ [:digit:]]{1,5}"))

#Taking the previously created vector and matching only digits. This should help me get rid characters I do not need and any other spaces.  

#Converting the pre score to numeric and unlisting it. 

pre_score <- unlist(as.numeric(str_extract_all(pre_score, "[[:digit:]]+")))



```

#Extracting Opponent Numbers
In this section, I need to extract the opponent numbers. As in the previous example, I find it more convenient to break this task into two steps. 

``` {r extract_opponent_id}
#In the first step, I want to extract the structure where a single upper case letter is followed by a space and a digit. The last part is quantified to occur at least two times, but not more than three times. 

opponent_id <- str_extract_all(mydata_names, "\\b [[:upper:] ][ [:digit:]]{2,3}")

#In the prior step, I was able to get part of the  output I needed, but it had extra spaces and characters that were not needed. In this step, I only match on digits to get rid of these extra characters.  

opponent_id <- str_extract_all(opponent_id, "[[:digit:]]{1,2}")

#I was new to the apply family of functions, but once I read about them it was what I needed for this task. In this next step, I am using the lapply to convert every element in the list to a numeric. This is needed for downstream analysis. 

opponent_id <- lapply(opponent_id, as.numeric)

head(opponent_id)

```

#Calculating Average Score

This is the fun part! I have to confess that it took me some time to figure it out since I was not familiar with these techniques in R.  

```{r calc_asp_opp}

# Creating a lookup function that will return the prescore based on the "i" index. This corresponds to the number of rows, so it should work in our example. 
lookup <- function (i){return (pre_score[i])}

#After creating the lookup function above, I then take advantage of the lapply function to match/replace player index ID and pre_scores. These values are then stored in the score list.

score <- lapply(opponent_id, lookup)

#At this point, I now have the pre_score values and can, once again, use lapply functions to calculate the avarege. for this, I need the sum of each of the score elements and the their length. Once I have those values, I can then take the average in the Average_score vector. 

pre_score_total <- unlist(lapply(score, sum))
pre_score_length <- unlist(lapply(score, length))
average_score <- round((pre_score_total)/(pre_score_length), 0)

#reviewing the average score, to see if at least the first row matches sample output. 
head(average_score)

#It does!
```

At last I have all the fields I need to create the requested output. This is done in the step below. I also gave the columns better looking names and added a column for the ID. This is simply a numeric vector with 64 elements. I also made  sure that my stringAsFactors parameter = FALSE. 

#Creating the Final Data Frame

``` {r fina_data_frame}

chess_df <- data.frame("ID" = 1:64, "Name"=mydata_names1, "State"=mydata_state, "Total Points"=mydata_total, "Pre-Rating"=pre_score,"Average Rating"=average_score, stringsAsFactors = FALSE)

head(chess_df)

```

#Writing the Final Output
In the last step, I will write the data frame to a csv file. This file will be written on my local working directory. 

``` {r writing_output}

write.csv(chess_df, file = "tournament_structure.csv")


```








