---
title: "Week_10_Assignment"
author: "Diego Diaz"
date: "November 1, 2015"
output: 
  html_document: 
    highlight: tango
    theme: readable
    toc: yes
---

#Assignment

The task is to choose one of the New York Times APIs, construct an interface in R to read in the JSON data, and
transform it to an R data frame.For this assignment, I chose the Movie Review API from the NYT's website. One drawback of this API is that paging is required. As a result, only 20 movies are returned at a time. 

#Required Libraries

```{r message=FALSE}
library(jsonlite)
library(dplyr)
```

#Retrieving Data from the NYT API

```{r get_data}
#API key from the NYT wesbsite for Movie Reviews 
key <- "&api-key=e13903a0a60761ade30cbf94df0f25cc:7:73345107"

#Request to access movie review from API. 
url <- "http://api.nytimes.com/svc/movies/v2/reviews/dvd-picks.json?order=by-title"

#Query API data from NYTimes website. The JSON function automatically 
query <- fromJSON(paste0(url, key))

```

#Subsetting Imported Data


```{r subsetting_dataframe}

reviews <- query$results
colnames(reviews)


```

> Keeping Columns of Interest Only

Using the **select** function from **dplyr** to keep only column of interest from the data frame. 

```{r columns}

#selecting desired columns
reviews_new <- select(reviews,nyt_movie_id, display_title, mpaa_rating, critics_pick, headline, summary_short,publication_date)

#column names
colnames(reviews_new)

#taking a peek
head(reviews_new)

```

#Structure of Imported Data


```{r structure_type}

#data structure to confirm it is a data frame.
str(reviews_new)


```

