---
title: "Assignment_Week_8"
author: "Diego Diaz"
date: "October 18, 2015"
output: 
  html_document: 
    highlight: tango
    theme: readable
    toc: yes
---

#Loading Required Libraries

```{r required_libraries}

library(XML)
library(rvest)
library(jsonlite)

```

#Project Objective and Data Structure

> Objective

The objective of the **Week 8 assignment** is to determine if data, produced in three different file conventions, will generate the same data frames once it is imported into **R**. The data file conventions are **HTML**, **XML**, **JSON**. 

These data sources have been posted to the following Github links: 

1. [books.html](https://raw.githubusercontent.com/diegomdiaz/IS607/master/Assignment%20Week%208/books.html) 
2. [books.xml](https://raw.githubusercontent.com/diegomdiaz/IS607/master/Assignment%20Week%208/books.xml)
3. [books.json](https://raw.githubusercontent.com/diegomdiaz/IS607/master/Assignment%20Week%208/books.json)


> Data Structure

In this assignment, the data is composed of attributes from three books. All three data sources have the same data and the following column names:

* Book_Title
* Author_Name
* Genre
* Pages

The book selections are as follow:

1. **The Martian** by Andy Wier. The genre is Science Fiction and it has 387 pages.

2. **The Hunt for Red October** by Tom Clancy. The genre is Thriller and Suspense and it has 387 pages. 

3. **Automated Data Collection with R** by Simon Munzert, Christian Rubba, Peter Meibner and Dominic Nyhuis.The genre is Computer Science and it has 480 pages

#HTML

> Importing the Data

In this example, I import the data from Git hub using the **html** function, from the **rvest** library, into the **url_html** container. 

```{r books_html_import}

#Importing the data from Github. 
url_html <- html("https://raw.githubusercontent.com/diegomdiaz/IS607/master/Assignment%20Week%208/books.html", encoding = "UTF-8")

```

> Converting to Data Frame

At this point, the data is now in the **url_html** container and I need to read the table within the data. For this task I use the **readHTMLTable** function to extract the table elements into the **books_html** variable. Interestingly, I am setting my **stringAsFactors** parameters = FALSE, but I still get factors in my data frame. 


```{r books_html_data_frame}

books_html <- readHTMLTable(url_html, header = TRUE, as.data.frame = TRUE, stringAsFactors = FALSE)

str(books_html)

```


#XML

> Importing the Data

In the prior HTML example, I imported the  data from Github. Here, I will work with an **XML** file saved to my working directory. For importing this file, I will use the **xmlParse** function from the **XML** library. One advantage of this function is that it "Validates" the **XML** document and will prompt if there are any issues with the file's DTD structure. 


```{r xml_import}
#using the Validate = True option to validate the file and DTD structure. 
books_xml <- xmlParse(file = "books.xml", validate = TRUE) 

#takign a peak at the imported xml code
str(books_xml)

```


> Converting to Data Frame

I now convert the data in the **books_xml** variable into a data frame using the **xmlToDataFrame** function from the **XML** library. 

```{r xml_converting2df}
dfxml <- xmlToDataFrame(books_xml, stringsAsFactors = FALSE)

str(dfxml)


```


#JSON

> Importing the Data & Converting to Data Frame

As in the XML example, I chose to import the **books.json** file from my working directory. I also selected the jsonlite library to both import the file into the **books_json** variable and, in the process, convert the file into a data frame. 

```{r json_import_and_dataframe}
#importing and convertign to data frame 
books_json <- fromJSON("books.json", simplifyDataFrame = TRUE)

#structure of the imported data frame
str(books_json)

```


#Conclusion

I was able to convert the books data sources into data frames in **R**. Once in a data frame, I was able to inspect the data using the **View()** function to generate a table. Additionally, I used the **str()** function to see the data structure for each data frame. 

Overall, all the file formats and conversion methods did a good job at turning the data into data frame; however, there were some subtle differences. In the **HTML** conversion, for example, the **readHTMLTable** function turned all the data values into factors even though I had **stringAsFactors = FALSE**. For the the **XML** conversion, the **stringAsFactors = FALSE** did work and the all the values were characters, including the pages. Finally, I was able to make the pages integers in the **JSON** file and did not need any additional coercion.   

In the **JSON** data frame, the Author Name with multiple authors was turned into a nested list. This did not displayed well in the table format. Lastly, the headers for each data frame were different as and varied as result of the file structure - the **XML** headers were the the clearest. 

I can conclude the resulting data frames were similar but not identical. 



